Lawvere-Tierney Sheafification\\ in Homotopy Type Theory

### 2015-12-11 paper/paper_acm.tex
reduce space v4

\usepackage{enumerate}
\usepackage{wasysym}
\usepackage{microtype}

@@

@@

Even if similar questions have been considered around the Curry-Howard
isomorphism, there is no canonical way to extend a programming
language such as type theory with new logical or computational
principle while keeping consistency automatically.
%
For instance, much efforts have been done to provide a computational
content to the law of excluded middle in order to define a
constructive version of classical logic. This has lead to various
calculi, with most notably the $\lambda \mu$-calculus of
Parigot~\cite{parigot1993classical}, but this line of work has not
appeared to be fruitful to define a new version of type theory with
classical principles.
%
Other works have tried to extend continuation-passing-style (CPS)
transformation to type theory, but they have been faced with the
difficulty that the CPS transformation is incompatible with (full) dependent
sums~\cite{barthe2002cps}, which puts emphasis on the tedious link
between the axiom of choice and the law of excluded middle in type theory.

@@

Using $\Type_n$ instead of just $\Type$ is the first step to connect
type theory to higher topos theory. The next step is to exhibit a
hierarchy of subobject classifiers on $n$-truncated homotopy fibers.
%
The homotopy fiber $\fib{f}{b}$ of a function $f$ at element $b$ is
defined as 
\begin{mymath}
\fib{f}{b} \defeq \sum_{a:A} f(a) = b.
\end{mymath} 
%
A function $f$ is with $n$-truncated homotopy fibers (or simply
$n$-truncated function) when $\fib{f}{b}$
is in $\Type_n$ for any $b$.  
%
Following~\cite{sets_in_hott}, it is possible to show that, for any
homotopy level $n$ and any type $B$, $\Type_n$ classifies subobjects
of $B$ with $n$-truncated homotopy fibers in the sense that there is
an equivalence
%
\begin{mymath}[0em]
  \chi : \sum_{A:\Type} \sum_{f:A \to B} \prod_{b\in B}
\IsType n\
\fib{f}{b} \xrightarrow{\sim} 
 (B \to \Type_n)
\end{mymath}
%
 such that the diagram
\begin{mymath}
\xymatrix{
  A \ar[r]^{\hspace{-1em} t_f} \ar[d]_f & \Type_n^\bullet \ar[d]^{\pi_1}\\
  B \ar[r]_{\hspace{-1em} \chi_f} & \Type_n
}
\end{mymath}
is a pullback for any $f$ with
 $n$-truncated homotopy fibers where $\Type_n^\bullet \defeq
 \sum_{A:\Type_n} A$ is the universe of pointed
$n$-truncated types and 
\begin{mymath}t_f = \lambda a,~(\fib{f}{f(a)},(a,\mathrm{idpath})).\end{mymath}

One has a way, from any type $T$, to construct a type
$\|T\|_n:\Type_n$ generated by
\begin{itemize}
\item a function $|\cdot|_n : T \to \|A\|_n$,
\item a proof of $\IsType n \|T\|_n$,
\end{itemize}
 satisfying the following universal property:
% Finally, we need to use mere propositions as basic elements of a logic
% when characterizing properties of types or functions. 
% %
% But as some constructors do not preserve the level of homotopy type
% (\eg sums or $\Sigma$-types), so we need to introduce $\squash{X}$, the
% {\em (propositional) truncation} of $X$ (also called bracket type, or
% squash type). This operation on types allows to truncate a type down
% to a mere proposition. Then for instance, the existential
% quantification on mere propositions ``there exists x:A such that
% P(x)'' can be defined as $ \squash{\sum_{x:A} P(x)}.  $

@@

\begin{defi}
  Let $f:A\to B$ be a function. The {\em image} of $f$ is defined as 
  \begin{mymath}\im(f) \defeq \sum_{y:B} \|\fib f y\|.\end{mymath}
\end{defi}

The canonical function $\hat f : A \to \im(f)$ is (-1)-connected,
being the left component of an orthogonal factorization system which
satisfies~\cite[Lemma 7.6.4]{hottbook}:
\begin{prop}
  A map $f:A\to B$ factors uniquely (up-to homotopy) through
  $\im(f)$ as a (-1)-connected function followed by a (-1)-truncated
  function.
\end{prop}

@@

\begin{proof}[Sketch of proof]
  By induction on truncation, we need to show that 
  \begin{mymath}\forall a,b:T_f,~\tilde f |a|_{n+1} = \tilde f |b|_{n+1} \simeq
  \modal (|a|_{n+1}=|b|_{n+1}).\end{mymath}

  We use the encode-decode~\cite[Section 8.9]{hottbook} method to characterize $\tilde f |a|_{n+1}
  = x$, and the result follows. We refer to the formalization for details.
\end{proof}

@@

\begin{mymath}\xymatrix{
\|X\|_{n+1} \ar[r] \ar[rd] & \|\mathring T_{X}\|_{n+1} \ar[r] \ar[d] & \|\mathring
  T_{\mathring T_X} \|_{n+1} \ar[ld]\cdots \\
  & Q &
} \end{mymath}

% \begin{mymath}\xymatrix @C=4em{ 
%     \cdots \overline{\Delta_3} \tar[r]{4pt} \ar[rd]_{f\circ \pi_1} & \overline{\Delta_2}
%     \dar[r]{2pt} \ar[d]^{f\circ \pi_1} &
%     \overline{\Delta_1} = T \ar[ld]^{f\circ \pi_1}\\
%     & Q &
%   }\end{mymath}
But we know (lemma~\ref{lem:sepiscolim}) that $\separated T$ is the
$(n+1)$-colimit of the diagram $\mathring T$, thus there is an universal
arrow $\separated T \to Q$.
%
This is enough to state the following proposition.
\begin{prop}\label{prop:sep-subu}
  $(\separated,\mu)$ defines a reflective subuniverse on $\Type_{n+1}$.
\end{prop}

@@

It remains to show that $\modal_{n+1}$ is left exact and is compatible
with $\modal_{-1}$. To do that, we need to extend the notion of
compatibility and show that actually every modality $\modal_{n+1}$ is
compatible with $\modal_n$ on lower homotopy types.
\begin{prop}
  If $T:\Type_n$, then $\modal_{n+1} \widehat T = \modal_n T$, where $\widehat T$ is $T$ seen as a
  $\Type_{n+1}$.
\end{prop}
\begin{proof}
  We prove it by induction on $n$:
  \begin{itemize}
  \item For $n=-1$: Let $T:\HProp$. Then
    \begin{mymath}
      \modal_{0} \widehat T = \sum_{u:T \to \Type_n^\modal} \modal_{-1}\left( \sum_{a:T} 
      (\lambda t,~\modal_{-1} (a=t)) = u\right)
    \end{mymath}
    because the type inside the truncation is already in $\HProp$.
    Now, let define $\phi : \modal_{-1} T \to \modal_0T$ by
    \begin{mymath}\phi t = (\lambda t',\, \modal_{-1}(t=t') ; \eta
    (t;\mathrm{idpath}))\end{mymath}
    and $\psi : \modal_0T\to \modal_{-1} T$ by obtaining the
    witness $a:T$ (which is possible because we are trying to inhabit
    a modal proposition), and letting
    \begin{mymath}\psi (u;x) = \eta_T a.\end{mymath}
    These two maps form an equivalence (the section and retraction are
    trivial because the equivalence is between mere propositions).
  \item Suppose now that $\modal_{n+1}$ is compatible with all $\modal_k$ on
    lower homotopy types. Let $\modal_{n+2}$ be as above, and let
    $T:\Type_{n+1}$. Then, as $\modal_{n+1}$ is compatible with $\modal_{n}$, and
    $(a=t)$ is in $\Type_n$,
    \begin{mymath}
      \modal_{n+2} \widehat T= \hspace{-1em} \sum_{u:T \to
        \Type_{n+1}^\modal} 
      \hspace{-1em} \modal_{-1}\left\| \sum_{a:T} 
        (\lambda t,~\modal_{n} (a=t)) = u\right\|.
    \end{mymath}

    It remains to prove that for every $(u,x)$ inhabiting the
    $\Sigma$-type above, $u$ is in $T\to\Type_n^\modal$, \ie that for
    every $t:T$, $\IsType n (u\, t)$.  But for any truncation index
    $p$,
    the type $\IsType p X:\HProp$ is a sheaf as soon as $X$ is, so we can get rid
    of $\modal_{-1}$ and of the truncation, which tells us that for
    every 
    $t:T$, $u\, t = \modal_n(a=t) : \Type_n$.
  \end{itemize}
\end{proof}

>>>>

\usepackage{enumerate}
\usepackage{wasysym}
\usepackage{microtype}
% \usepackage[subtle]{savetrees}

@@

\let\oldthebibliography=\thebibliography
  \let\endoldthebibliography=\endthebibliography
  \renewenvironment{thebibliography}[1]{%
    \begin{oldthebibliography}{#1}%
      \setlength{\parskip}{0ex}%
      \setlength{\itemsep}{0ex}%
  }%
  {%
    \end{oldthebibliography}%
  }

@@

Even if similar questions have been considered around the Curry-Howard
isomorphism, there is no canonical way to extend a programming
language such as type theory with new logical or computational
principle while keeping consistency automatically.
%
For instance, much efforts have been done to provide a computational
content to the law of excluded middle in order to define a
constructive version of classical logic. This has lead to various
calculi, with most notably the $\lambda \mu$-calculus of
Parigot~\cite{parigot1993classical}, but this line of work has not
appeared to be fruitful to define a new version of type theory with
classical principles.
%
Other works have tried to extend continuation-passing-style (CPS)
transformation to type theory, but they have been faced with the
difficulty that the CPS transformation is incompatible with (full) dependent
sums~\cite{barthe2002cps}, which puts emphasis on the tedious link
between the axiom of choice and the law of excluded middle in type theory\looseness=-1.

@@

Using $\Type_n$ instead of just $\Type$ is the first step to connect
type theory to higher topos theory. The next step is to exhibit a
hierarchy of subobject classifiers on $n$-truncated homotopy fibers.
%
The homotopy fiber $\fib{f}{b}$ of a function $f$ at element $b$ is
defined as 
\begin{mymath}
\fib{f}{b} \defeq \sum_{a:A} f(a) = b.
\end{mymath} 
%
A function $f$ is with $n$-truncated homotopy fibers (or simply
$n$-truncated function) when $\fib{f}{b}$
is in $\Type_n$ for any $b$.  
%
Following~\cite{sets_in_hott}, it is possible to show that, for any
homotopy level $n$ and any type $B$, $\Type_n$ classifies subobjects
of $B$ with $n$-truncated homotopy fibers in the sense that there is
an equivalence
%
\begin{mymath}[0em]
  \chi : \sum_{A:\Type} \sum_{f:A \to B} \prod_{b\in B}
\IsType n\
\fib{f}{b} \xrightarrow{\sim} 
 (B \to \Type_n)
\end{mymath}
%
 such that the diagram
\begin{mymath}
\xymatrix{
  A \ar[r]^{\hspace{-1em} t_f} \ar[d]_f & \Type_n^\bullet \ar[d]^{\pi_1}\\
  B \ar[r]_{\hspace{-1em} \chi_f} & \Type_n
}
\end{mymath}
is a pullback for any $f$ with
 $n$-truncated homotopy fibers where $\Type_n^\bullet \defeq
 \sum_{A:\Type_n} A$ is the universe of pointed
$n$-truncated types and 
\begin{mymath}t_f = \lambda a,~(\fib{f}{f(a)},(a,\mathrm{idpath})).\end{mymath}
One has a way, from any type $T$, to construct a type
$\|T\|_n:\Type_n$ generated by
\begin{itemize}
\item a function $|\cdot|_n : T \to \|A\|_n$,
\item a proof of $\IsType n \|T\|_n$,
\end{itemize}
 satisfying the following universal property:
% Finally, we need to use mere propositions as basic elements of a logic
% when characterizing properties of types or functions. 
% %
% But as some constructors do not preserve the level of homotopy type
% (\eg sums or $\Sigma$-types), so we need to introduce $\squash{X}$, the
% {\em (propositional) truncation} of $X$ (also called bracket type, or
% squash type). This operation on types allows to truncate a type down
% to a mere proposition. Then for instance, the existential
% quantification on mere propositions ``there exists x:A such that
% P(x)'' can be defined as $ \squash{\sum_{x:A} P(x)}.  $

@@

\begin{defi}
  Let $f:A\to B$ be a function. The {\em image} of $f$ is defined as 
  \begin{mymath}\im(f) \defeq \sum_{y:B} \|\fib f y\|.\end{mymath}
\end{defi}
The canonical function $\hat f : A \to \im(f)$ is (-1)-connected,
being the left component of an orthogonal factorization system which
satisfies~\cite[Lemma 7.6.4]{hottbook}:
\begin{prop}
  A map $f:A\to B$ factors uniquely (up-to homotopy) through
  $\im(f)$ as a (-1)-connected function followed by a (-1)-truncated
  function.
\end{prop}

@@

\begin{proof}[Sketch of proof]
  By induction on truncation, we need to show that 
  \begin{mymath}\forall a,b:T_f,~\tilde f |a|_{n+1} = \tilde f |b|_{n+1} \simeq
  \modal (|a|_{n+1}=|b|_{n+1}).\end{mymath}
  We use the encode-decode~\cite[Section 8.9]{hottbook} method to characterize $\tilde f |a|_{n+1}
  = x$, and the result follows. We refer to the formalization for details.
\end{proof}

@@

\begin{mymath}\xymatrix{
\|X\|_{n+1} \ar[r] \ar[rd] & \|\mathring T_{X}\|_{n+1} \ar[r] \ar[d] & \|\mathring
  T_{\mathring T_X} \|_{n+1} \ar[ld]\cdots \\
  & Q &
} \end{mymath} 
% \begin{mymath}\xymatrix @C=4em{ 
%     \cdots \overline{\Delta_3} \tar[r]{4pt} \ar[rd]_{f\circ \pi_1} & \overline{\Delta_2}
%     \dar[r]{2pt} \ar[d]^{f\circ \pi_1} &
%     \overline{\Delta_1} = T \ar[ld]^{f\circ \pi_1}\\
%     & Q &
%   }\end{mymath}
But we know (lemma~\ref{lem:sepiscolim}) that $\separated T$ is the
$(n+1)$-colimit of the diagram $\mathring T$, thus there is an universal
arrow $\separated T \to Q$.
%
This is enough to state the following proposition.
\begin{prop}\label{prop:sep-subu}
  $(\separated,\mu)$ defines a reflective subuniverse on $\Type_{n+1}$.
\end{prop}

@@

It remains to show that $\modal_{n+1}$ is left exact and is compatible
with $\modal_{-1}$. To do that, we need to extend the notion of
compatibility and show that actually every modality $\modal_{n+1}$ is
compatible with $\modal_n$ on lower homotopy types.
\begin{prop}
  If $T:\Type_n$, then $\modal_{n+1} \widehat T = \modal_n T$, where $\widehat T$ is $T$ seen as a
  $\Type_{n+1}$.
\end{prop}
\begin{proof}
  We prove it by induction on $n$:
  \begin{itemize}
  \item For $n=-1$: Let $T:\HProp$. Then
    \begin{mymath}
      \modal_{0} \widehat T = \sum_{u:T \to \Type_n^\modal} \modal_{-1}\left( \sum_{a:T} 
      (\lambda t,~\modal_{-1} (a=t)) = u\right)
    \end{mymath}
    because the type inside the truncation is already in $\HProp$.
    Now, let define $\phi : \modal_{-1} T \to \modal_0T$ by
    \begin{mymath}\phi t = (\lambda t',\, \modal_{-1}(t=t') ; \eta
    (t;\mathrm{idpath}))\end{mymath}
    and $\psi : \modal_0T\to \modal_{-1} T$ by obtaining the
    witness $a:T$ (which is possible because we are trying to inhabit
    a modal proposition), and letting
    \begin{mymath}\psi (u;x) = \eta_T a.\end{mymath}
    These two maps form an equivalence (the section and retraction are
    trivial because the equivalence is between mere propositions).
  \item Suppose now that $\modal_{n+1}$ is compatible with all $\modal_k$ on
    lower homotopy types. Let $\modal_{n+2}$ be as above, and let
    $T:\Type_{n+1}$. Then, as $\modal_{n+1}$ is compatible with $\modal_{n}$, and
    $(a=t)$ is in $\Type_n$,
    \begin{mymath}
      \modal_{n+2} \widehat T= \hspace{-1em} \sum_{u:T \to
        \Type_{n+1}^\modal} 
      \hspace{-1em} \modal_{-1}\left\| \sum_{a:T} 
        (\lambda t,~\modal_{n} (a=t)) = u\right\|.
    \end{mymath}
    It remains to prove that for every $(u,x)$ inhabiting the
    $\Sigma$-type above, $u$ is in $T\to\Type_n^\modal$, \ie that for
    every $t:T$, $\IsType n (u\, t)$.  But for any truncation index
    $p$,
    the type $\IsType p X:\HProp$ is a sheaf as soon as $X$ is, so we can get rid
    of $\modal_{-1}$ and of the truncation, which tells us that for
    every 
    $t:T$, $u\, t = \modal_n(a=t) : \Type_n$.
  \end{itemize}
\end{proof}
